Dependently Typed Languages in Statix

### 2022-10-13 paper-evcs/oasics-v2021-sample-article.tex
Rename scope relations + work on paper

@@

\subsection{How scope graphs are used}

@@

Both of these need to be stored in the scope graph. The scope graph only has a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NameType|, which stores the type of a name, or a \verb|NameSubst|, which stores a substitution corresponding to a name.

\begin{lstlisting}
signature sorts	NameEntry
constructors    NameType : Expr -> NameEntry
                NameSubst : scope * Expr -> NameEntry
relations       name : ID -> NameEntry
name-resolution	labels P
\end{lstlisting}

@@

\begin{lstlisting}
scopePutType  : scope * ID * Expr -> scope
scopePutSubst : scope * ID * (scope * Expr) -> scope
scopeGetName  : scope * ID -> NameEntry
scopeGetNames : scope * ID -> list((path * (ID * NameEntry)))
empty_scope   : -> scope
\end{lstlisting}

The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively. To query the scope graph, use \verb|scopeGetName| or \verb|scopeGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found. Finally, \verb|empty_scope| returns a fresh empty scope.

@@

A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare. Beta reduction is done using Krivine abstract machines\cite{krivine}. We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications. The scope acts as the environment from \cite{krivine} paper, using \verb|NameSubst| to store substitutions. Furthermore, we can define \verb|expectBetaEq|, which asserts that two terms are equal. This calls \verb|expectBetaEq_| with \verb|betaReduceHead| applied.

@@

>>>>

\usepackage{mathpartir}

@@

\subsection{Scope Graphs}

@@

Both of these need to be stored in the scope graph. The scope graph only has a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a substitution corresponding to a name.

\begin{lstlisting}
signature sorts	NameEntry
constructors    NType : Expr -> NameEntry
                NSubst : scope * Expr -> NameEntry
relations       name : ID -> NameEntry
name-resolution	labels P
\end{lstlisting}

@@

\begin{lstlisting}
sPutType  : scope * ID * Expr -> scope
sPutSubst : scope * ID * (scope * Expr) -> scope
sGetName  : scope * ID -> NameEntry
sGetNames : scope * ID -> list((path * (ID * NameEntry)))
sEmpty    : -> scope
\end{lstlisting}

The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively. To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found. Finally, \verb|sEmpty| returns a fresh empty scope.

@@

A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare. Beta reduction is done using Krivine abstract machines\cite{krivine}. We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications. The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions. Furthermore, we can define \verb|expectBetaEq|, which asserts that two terms are equal. This calls \verb|expectBetaEq_| with \verb|betaReduceHead| applied.

@@

We can then start defining concepts in this language. First, we present the inference rules of the language:

\begin{figure}

\begin{mathpar}
\inferrule{ }{s \vdash Type() : Type()}

\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }

\inferrule{ sGetName(s, n) = NType(t) }{ s \vdash Var(n) : t }

\inferrule{ sGetName(s, n) = NSubst(se, e) \and se \vdash e : t }{ s \vdash Var(n) : t }

\inferrule{ s \vdash a : at \and at \underset{\beta}{=} Type() \and a \underset{\beta}{\Rightarrow} a' \and sPutType(s, n, a') \vdash b : bt \and bt \underset{\beta}{=} Type() } { s \vdash FnType(n, a, b) : Type() }

\inferrule{ s \vdash a : at \and at \underset{\beta}{=} Type() \and a \underset{\beta}{\Rightarrow} a' \and sPutType(s, n, a') \vdash b : bt } { s \vdash FnConstruct(n, a, b) : FnType(n, a', bt) }

\end{mathpar}
\caption{Rules for type-checking the Calculus of Constructions}
\label{fig:type-check-rules}
\end{figure}

\begin{figure}
\begin{mathpar}
\inferrule{ }{ s, Type(), [] \Rightarrow s, Type() }

\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
\end{mathpar}
\caption{Rules for beta reducing the Calculus of Constructions}
\label{fig:beta-head-rules}
\end{figure}
